Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-19 index.qmd
section:background,structural; process:write

<!-- 1. What does empirical research do first? -->
Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation. In this context, researchers typically frame their research questions as *estimands* or *targets of inference*. These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010]. For instance, a study might examine the question, “To what extent do different teaching methods ($T$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(T_{i}=\{1,2\})$. Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y_{i}=\{0,1\})$, indicating which student among those compared has a higher level of understanding. The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed by the associational quantity $E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, where $E[\cdot]$ denotes the expected value. An example of this approach is seen in @Jones_et_al_2019.

<!-- 2. What do they do next? -->
Researchers then proceed to identify the estimands. *Identification* refers to the process of accurately computing an estimand using an estimator. An *estimator* is a method or function that maps data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refer to the process of integrating data with an estimator [@Everitt_et_al_2010]. Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates. For instance, the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the target of inference discussed earlier, as long as the assumptions of the model are met.

<!-- 3. But we want these estimands to be causal -->
However, many studies aim to understand the mechanisms underlying specific data types and also seek to establish causal relationships rather than merely infer associations. In the earlier example, the differences between groups obtained using the BTL model, referred to as the associational estimates, can be interpreted as causal because the data were collected through a randomized experiment. Randomized experiments enable the causal interpretation of associational estimates by ensuring several key properties, including common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020]. The most crucial property, however, is that randomization effectively eliminates confounding. *Confounding* occurs when an external variable influences both the outcome and the variable of interest, leading to spurious associations [@Everitt_et_al_2010]. Randomization mitigates this issue by decoupling the intervention assignment mechanism, such as assigning students to different groups, from other variables and outcomes [@Morgan_et_al_2014; @Neal_2020].

<!-- 4. Experiments are not available to everyone, then causal inference comes to the rescue -->
Experiments are widely regarded as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014], however, researchers often encounter constraints that limit their ability to conduct experimental studies. These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020]. In these instances, causal inference offers a valuable alternative for generating causal estimates. *Causal inference* is a framework designed to identify the causes of phenomena and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, causal inference provides a rigorous mathematical framework for analyzing causes and counterfactuals [@Pearl_2009]. Consequently, the framework offers significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].

<!-- 6. But how does it work? -->
While a detailed explanation of causes and counterfactuals is beyond the scope of this document, it is important to note that causal inference addresses the *fundamental problem of causality* through the use of counterfactuals [@Neal_2020]. The framework defines *individual causal effects* (ICE) as the difference between an individual's observed and unobserved potential outcomes. These unobserved potential outcomes, known as *counterfactuals*, represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes for an individual, resulting from a specific cause, are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. The framework then extends this estimand to *average causal effects* (ACE), which captures the difference between sample averages of observed potential outcomes and counterfactuals. Finally, much like in an experiment, the framework identifies the ACE from associational estimates by conditioning for a *sufficient adjustment set* of variables, ensuring the absence of confounding [@Morgan_et_al_2014].

@@

<!-- 8. But which approach are we going to use? -->
The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of theories through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]. DAGs function as heuristics, effectively conveying the presumed causal structure of a system, referred to as a *scientific model*. They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020]. The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents this process of transitioning from estimands to estimates, as well as the application of the scientific model and data to identify and estimate causal effects.

>>>>

<!-- 1. What does empirical research do first? -->
Empirical research addresses real-world challenges by relying on evidence gathered through observation and experimentation. In this context, researchers typically frame their research questions as *estimands* or *targets of inference*. These estimands represent the specific quantities the study aims to determine [@Everitt_et_al_2010]. For instance, a study might examine the question, “To what extent do different teaching methods ($T$) influence students’ conceptual understanding of a topic ($Y$)?" To investigate this, the study could randomly assign students to two groups, each using a different teaching method $(T=\{1,2\})$. Students’ conceptual understanding of the topic could be evaluated through pairwise comparisons, resulting in a dichotomous outcome $(Y=\{0,1\})$, indicating which student among those compared has a higher level of understanding. The research question could be then framed as the estimand, “*On average*, is there a difference in conceptual understanding of the topic between the two groups of students?” This estimand could be mathematically expressed by the associational random quantity $E[Y| T=1] - E[Y| T=2]$, where $E[\cdot]$ denotes the expected value. An example of this approach is seen in @Jones_et_al_2019.

<!-- 2. What do they do next? -->
Researchers then proceed to identify the estimands. *Identification* refers to the process of accurately computing an estimand using an estimator. An *estimator* is a method or function that maps data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand and are derived through *estimation*, which refers to the process of integrating data with an estimator [@Everitt_et_al_2010]. Although various methods can approximate an estimand, researchers prioritize estimators with desirable properties that ensure the accuracy of estimates. For instance, the Z-test is an estimator known for its effectiveness in comparing two proportions, yielding accurate estimates when its underlying assumptions are met [@Kanji_2006]. The Z-test is expressed as a signal-to-noise ratio: $Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}$. The signal is the difference between the group sample proportions, $\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}$ and $\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}$, analogous to $E[Y| T=1]$ and $E[Y| T=2]$, respectively. The noise $\hat{s}_{p}$ is the unpooled sample variability observed between the two groups.

<!-- the BTL model [@Bradley_et_al_1952; @Luce_1959] is an estimator known for its effectiveness in modeling pairwise comparisons, yielding accurate estimates for the comparison of two groups, as long as the assumptions of the model are met. Usually these estimates are expressed in terms of the proportions of groups  -->

<!-- 3. But we want these estimands to be causal -->
However, many studies aim to understand the mechanisms underlying specific data and also seek to establish causal relationships rather than merely infer associations. In the earlier example, the differences between groups obtained using the Z-test, referred to as the associational estimate, can be interpreted as causal because the data were collected through a randomized experiment. Randomized experiments enable the causal interpretation of associational estimates by ensuring several key properties, including common support, no interference, and consistency [@Morgan_et_al_2014; @Neal_2020]. The most crucial property, however, is that randomization effectively eliminates confounding. *Confounding* occurs when an external variable influences both the outcome and the variable of interest, leading to spurious associations [@Everitt_et_al_2010]. Randomization mitigates this issue by decoupling the intervention assignment mechanism, such as assigning students to different groups, from other variables and outcomes [@Morgan_et_al_2014; @Neal_2020].

<!-- 4. Experiments are not available to everyone, then causal inference comes to the rescue -->
Experiments are widely recognized as the gold standard in evidence-based science [@Hariton_et_al_2018; @Hansson_2014]. However, researchers often face constraints that limit their ability to conduct experimental studies. These constraints include ethical concerns, such as the assignment of individuals to potentially harmful interventions, and practical limitations, such as the infeasibility of, for example, assigning individuals to genetic modifications or physical impairments [@Neal_2020]. In these situations, causal inference provides a valuable alternative for generating causal estimates, particularly when the goal is to understand the mechanisms underlying specific data. Moreover, the framework offers significant theoretical insights that enhance the design of observational and experimental studies [@McElreath_2020].

<!-- 5. What is causal inference? -->
Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, *causal inference* is a framework designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. This framework employs rigorous mathematical techniques to address the fundamental problem of causality [@Pearl_2009], which is essential for understanding and defining causal effects. The *fundamental problem of causality* centers on answering the question, "What would have happened 'in the world' under different circumstances?" a concept known as counterfactuals. *Counterfactuals* represent hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a specific cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024].

<!-- 6. But how does it work? --> 
Although a comprehensive discussion of causes and counterfactuals exceeds the scope of this document, a brief overview of how the framework addresses the fundamental problem of causality is possible using the previous example. The framework starts by defining *individual causal effects* (ICE) as the difference between students' observed and unobserved potential outcomes: $\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)$. Notice once a student is assigned to $T_{i}=1$, the potential outcome under $T_{i}=2$ is no longer observed nor observable, and is thus termed a *counterfactual*. To address the challenge posed by counterfactuals, the framework extends the ICE to *average causal effects* (ACE). The ACE is defined as $\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]$, and represents the difference between the average of observed potential outcomes and counterfactuals in the sample. Finally, akin to experimental studies, the ACE is identified from associational estimates by ensuring the absence of confounding. This is accomplished by statistically conditioning on a *sufficient adjustment set* of variables ($X$). Consequently, the ACE is expressed as $\tau = E_{X}[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]]$, where $E_{X}[\cdot]$ denotes the marginal expected value over $X$ [@Morgan_et_al_2014].

@@

<!-- 8. But which approach are we going to use? -->
The structural approach, however, offers a notable advantage over the potential outcomes approach by allowing the graphical and formal representation of the ACE through directed acyclic graphs [DAG, @Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020]. DAGs function as heuristics, effectively conveying the presumed causal structure of a system, referred to as a *scientific model*. They do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020]. The Identification-Estimation flowchart in @fig-IEflow [@McElreath_2020; @Neal_2020] visually represents the process of transitioning from estimands to estimates, as well as the application of the scientific model and data to identify and estimate causal effects.
